Distortion-corrected kernel density estimate on Riemannian manifolds

### 2022-09-08 paper/paper.Rmd
Minor

---
title: "Distortion-corrected kernel density estimate on Riemannian manifolds"
author:
- familyname: Cheng
  othernames: Fan
  address: Monash University
  email: Fan.Cheng@monash.edu
  # correspondingauthor: true
- familyname: Panagiotelis
  othernames: Anastasios
  address: University of Sydney
  email: Anastasios.Panagiotelis@sydney.edu.au
- familyname: Hyndman
  othernames: Rob J
  address: Monash University
  email: Rob.Hyndman@monash.edu
abstract: "Manifold learning can be used to obtain a low-dimensional representation of the underlying Riemannian manifold given the high-dimensional data. However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the Riemannian manifold. We propose a novel distortion-corrected kernel density estimator (DC-KDE) for any manifold learning embedding by introducing the estimated Riemannian metric of each point to fix the distortion in the line and volume elements. The geometric information of the manifold guarantees a more accurate density estimation of the true manifold, which subsequently could be used for anomaly detection. To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations with a 2-D data from Gaussian mixture model mapped into a 3-D twin peaks shape and a 5-D semi-hypersphere mapped in a 100-D space. We demonstrate that the proposed DC-KDE could improve the density estimates given a good manifold learning embedding and has higher rank correlations with the true manifold density. A shiny app in R is also developed for various simulation scenarios. The proposed method is applied to density estimation in statistical manifolds of electricity usage with the Irish smart meter data. This demonstrates the estimator's capability to fix the distortion of the manifold geometry and a new approach to anomaly detection in high-dimensional data."
keywords: "manifold learning, variable bandwidth, Riemannian metric, geodesic distance, Gaussian kernels"
# wpnumber: 03/2021
# jelcodes: C55, C65, C80
blind: false
cover: false
toc: false
lot: false
bibliography: references.bib
biblio-style: authoryear-comp
output:
  monash::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
      # after_body: appendix.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
    latex_engine: pdflatex
editor_options: 
  markdown: 
    wrap: 72
---

@@

<!-- embedding pushforward metric [@Perrault-Joncas2013-pq] -->

<!-- ## Volume density function [MOVED TO SECTION 2.2]

For any tangent vector $u \in T_pM$, let $\gamma(u): t \rightarrow \exp_p(\frac{tu}{\|u\|})$ and let $w_i, i=1,\dots,d$ be Jacobi fields along $\gamma$ such that $w_i(0)=0$, for all $i=1,\dots,d$, $Dw_1/dt(0)=u/\|u\|$. Then $Dw_i/dt(0)$ forms an orthonomal basis of the tangent space $T_pM$. Define the volume density function on $M$, $\theta_p(q): T_pM \rightarrow \mathbb{R}$, as 
$$
\theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
$$

The exponential map $\exp_p: T_pM \rightarrow M$ induced by pullback of a volume form $\exp_p^*\textit{Vol}$ on $T_pM$, and $\theta_p$ is its density with respect to the Lebesgue measure of the Euclidean structure on $T_pM$, which gives in normal coordinates 
$$
d\exp_p^*\textit{Vo}l(u) = \theta_p(u)du.
$$
The volume density function is motivated by how a measure $du$ changes under the change of coordinates. It is related to the integration on manifolds and plays an important part in studing the density of manifolds. -->

<!-- the volume density function is the quotient of the canonical measure of the Riemannian metric $\exp_p{g}$ on $T_pM$ by the Lebesgue measure of the Euclidean structure $g_p$ on $T_pM$. -->

<!-- ## Example of $S^1$ for manifold density estimation [TODO: rewrite as appendix B with plots?] -->

<!-- Manifold: $S^1$ with center 0 and radius $r=1$ -->

<!-- Polar coordinates: $(r, \theta)$ -->

<!-- Uniformly distributed angle: $\theta \sim U(0, \pi/2)$ with $f(\theta) = 2 / \pi$ for $\theta \in [0, \pi /2]$. -->

<!-- Cartesian coordinates: $(x, y)$ where $x = \cos(\theta) \in [0, 1]$ and $y = \sin(\theta) \in [0, 1]$ -->

<!-- Density of $\theta$: $f(\theta) = 2 / \pi$ -->

<!-- Density of $x$: $f(x) = \frac{2}{\pi \sqrt{1-x^2}}$ -->

<!-- The CDF of X is given by  -->
<!-- $$ -->
<!-- P(X \leq x) = P(\cos(\theta) \leq x) \\ -->
<!-- = P(\theta \geq \arccos(x)) \\ -->
<!-- = \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\ -->
<!-- =  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\ -->
<!-- = 1 - \frac{2}{\pi} \arccos(x). -->
<!-- $$  -->
<!-- Then the density of X is the derivative of the CDF w.r.t. $x$ $$ -->
<!-- f(x) = -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\ -->
<!-- = \frac{2}{\pi \sqrt{1 - x^2}}. -->
<!-- $$ -->

<!-- Given $(x, y) \in \mathbb{R}^2$ as sample points from $S^1$, we could -->
<!-- estimate the density of each sample point $X_i \in S^1, i = 1, \dots, N$ -->
<!-- and compare it with its true density $f(x)$. -->
<!-- We use the density estimator in @Pelletier2005-vu. For point $p \in S^1$, the density at $p$ is estimated by -->
<!-- $$ -->
<!-- f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{r}), -->
<!-- $$  -->
<!-- where $d=1$, $r$ is the bandwidth, $\theta_{X_i}(p)$ is the volume density function at sample point $X_i$, and $d_g(p, X_i)$ is the geodesic distance between $p$ and $X_i$ along the manifold $S^1$. -->

<!-- Since the manifold is known ($S^1$), we know the geodesic distance is the arc length between two points on the manifold, which is given by $L(\gamma) = r(\beta - \alpha)$ with $r=1$. Therefore, the geodesic distance is  -->
<!-- $$ -->
<!-- d_g(p, X_i) = \arccos(p) - \arccos(X_i) -->
<!-- $$  -->
<!-- As stated in @Pelletier2005-vu, in terms of *geodesic normal coordinates* at $X_i$, the volume density function $\theta_{X_i}(p)$ equals the determinant of the metric $g$ expressed in these coordinates at the logarithm map $\exp_{X_i}^{-1}(p)$, i.e. $|g(\exp_p^{-1}(x_i))|$. -->

<!-- [When using normal coordinates(orthogonal basis) for tangent vector in $T_{X_i}S^1$, the exponential map has *geodesic normal coordinates*.] -->

<!-- The Riemannian metric $g$ for $S^1$ is given by the inner products on the tangent spaces of $S^1$. For each $X_i \in M$,  -->
<!-- $$ -->
<!-- g_{X_i}: T_{X_i}M \times T_{X_i}M \rightarrow \pmb{R}, -->
<!-- $$  -->
<!-- is smooth in a local neighborhood $U$ of $M$. -->

<!-- Then the volume density function $\theta_{X_i}(p)$ is the inner product of the tangent vectors produced by the logarithm map from $X_i$ to $p$, which is given by  -->
<!-- $$ -->
<!-- \theta_{X_i}(p) = \sqrt{|\cos(d_g(p,q))|} = \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}. -->
<!-- $$  -->
<!-- Therefore, the estimator for $p \in S^1$ is given by  -->
<!-- $$ -->
<!-- f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}} K\bigg(\frac{\|\arccos(p) - \arccos(X_i))\|}{r}\bigg), -->
<!-- $$  -->
<!-- where the bandwidth $r$ is selected with the plug-in bandwidth selector implemented by the R package *ks*. -->

<!-- ### How to derive volume density function $\theta_{X_i}(p)$ -->

<!-- #### Example of $S^2$ in @Henry2009-om. -->

<!-- Let $S^2$ be the two-dimensional sphere of radius 1 and $p \in S^2$. Let $v$ and $w$ be vectors such that $\|v\| = \|w\| = 1$ and $\{v, w, p\}$ is the orthonormal basis of $\mathbb{R}^3$. Consider the exponential chart $(U, \Psi)$ induced by the parametrisation $\Psi^{-1}: B_\pi(0) \rightarrow S^2 \backslash \{-p\}$ given by  -->
<!-- $$ -->
<!-- \Psi^{-1}(s,t) = \cos(r)p + \frac{\sin(r)}{r}(sv+tw), -->
<!-- $$  -->
<!-- if $(s,t) \neq (0,0)$ and $\Psi^{-1}(0,0)=p$ where $r=\sqrt{s^2 + t^2}$.  -->
<!-- Note that $r=d_g(p,q)$ if $q = \exp_p(sv+tw)$. -->
<!-- Then we take the partial derivatives of $\Psi^{-1}(s,t)$ with respect to the basis of the tangent space, $s$ and $t$, and have that -->
<!-- $$ -->
<!-- A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}s}{r}p + \frac{\cos(r)rs^2+t^2\sin(r)}{r^3}v + \frac{(\cos(r)r-\sin(r))st}{r^3}w, -->
<!-- $$ -->
<!-- $$ -->
<!-- B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}t}{r}p + \frac{(\cos(r)r-\sin(r))st}{r^3}v + \frac{\cos(r)rt^2+\sin(r)s^2}{r^3}w. -->
<!-- $$  -->
<!-- The coefficients for $v$ and $w$ gives the local coordinates of two tangent vectors. -->
<!-- Further, since the riemannian metric is defined as the inner product between two tangent vectors, each component of the riemannian metric could be calculated as $g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))$.  -->
<!-- Therefore, we obtain that the volume density on the sphere is  -->
<!-- $$ -->
<!-- \theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2} \\ -->
<!-- = \frac{|\sin(d_g(p,q))|}{d_g(p,g)}, -->
<!-- $$  -->
<!-- for $q \neq p, -p$, and $\theta_p(p) = 1$. -->

<!-- #### $S^1$ -->

<!-- Similarly in $S^1$, we could let $\{v, p\}$ be the orthonormal basis of $\mathbb{R}^2$. Then the exponential chart is induced by the parametrization  -->
<!-- $$ -->
<!-- \Psi^{-1}(s) = \cos(r)p + \frac{\sin(r)}{r}sv, -->
<!-- $$ -->
<!-- and $\Psi^{-1}(0) = p$ where $r = s$. Again $r = d_g(p,q)$ if $q = \exp_p(sv)$. Then we have -->
<!-- $$ -->
<!-- A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = - \sin(r)p + \cos(r)v. -->
<!-- $$ -->
<!-- So the riemannian metric is $g_q = \cos(r)$ and further,  -->
<!-- $$ -->
<!-- \theta_p(q) = \sqrt{|\cos(r)|} = \sqrt{|\cos(d_g(p,q))|}, -->
<!-- $$ -->
<!-- where the geodesic distance is $d_g(p,q) = \arccos(p) - \arccos(q)$, and $\theta_p(p) = 1$. -->

>>>>

---
title: "Distortion-corrected kernel density estimate on Riemannian manifolds"
author:
- familyname: Cheng
  othernames: Fan
  address: Monash University
  email: Fan.Cheng@monash.edu
  # correspondingauthor: true
- familyname: Panagiotelis
  othernames: Anastasios
  address: The University of Sydney
  email: Anastasios.Panagiotelis@sydney.edu.au
- familyname: Hyndman
  othernames: Rob J
  address: Monash University
  email: Rob.Hyndman@monash.edu
abstract: "Manifold learning can be used to obtain a low-dimensional representation of the underlying Riemannian manifold given the high-dimensional data. However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the Riemannian manifold. We propose a novel distortion-corrected kernel density estimator (DC-KDE) for any manifold learning embedding by introducing the estimated Riemannian metric of each point to fix the distortion in the line and volume elements. The geometric information of the manifold guarantees a more accurate density estimation of the true manifold, which subsequently could be used for anomaly detection. To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations with a 2-D data from Gaussian mixture model mapped into a 3-D twin peaks shape and a 5-D semi-hypersphere mapped in a 100-D space. We demonstrate that the proposed DC-KDE could improve the density estimates given a good manifold learning embedding and has higher rank correlations with the true manifold density. A shiny app in R is also developed for various simulation scenarios. The proposed method is applied to density estimation in statistical manifolds of electricity usage with the Irish smart meter data. This demonstrates the estimator's capability to fix the distortion of the manifold geometry and a new approach to anomaly detection in high-dimensional data."
keywords: "manifold learning, variable bandwidth, Riemannian metric, geodesic distance, Gaussian kernels"
# wpnumber: 03/2021
# jelcodes: C55, C65, C80
blind: false
cover: false
toc: false
lot: false
bibliography: references.bib
biblio-style: authoryear-comp
output:
  monash::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
      # after_body: appendix.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
    latex_engine: pdflatex
editor_options: 
  markdown: 
    wrap: 72
---

@@

# Appendix: Simulation of $S^1$ for manifold density estimation {#s1simulation}

To start with, we demonstrate the use of the proposed DC-KDE on a $1$-dimensional example. For a $S^1$ curve centered at $O$ with radius $1$, the coordinated of points on the curve could be expressed in polar coordinates $(1, \alpha)$ or Cartesian coordinates $(x, \sqrt{1-x^2})$.

The first simulation starts with a $1$-dimensional example.

Manifold: $S^1$ with center 0 and radius $r=1$

Polar coordinates: $(r, \theta)$

Uniformly distributed angle: $\theta \sim U(0, \pi/2)$ with $f(\theta) = 2 / \pi$ for $\theta \in [0, \pi /2]$.

Cartesian coordinates: $(x, y)$ where $x = \cos(\theta) \in [0, 1]$ and $y = \sin(\theta) \in [0, 1]$

Density of $\theta$: $f(\theta) = 2 / \pi$

Density of $x$: $f(x) = \frac{2}{\pi \sqrt{1-x^2}}$

The CDF of X is given by
$$
P(X \leq x) = P(\cos(\theta) \leq x) \\
= P(\theta \geq \arccos(x)) \\
= \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\
=  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\
= 1 - \frac{2}{\pi} \arccos(x).
$$
Then the density of X is the derivative of the CDF w.r.t. $x$ $$
f(x) = -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\
= \frac{2}{\pi \sqrt{1 - x^2}}.
$$

Given $(x, y) \in \mathbb{R}^2$ as sample points from $S^1$, we could
estimate the density of each sample point $X_i \in S^1, i = 1, \dots, N$
and compare it with its true density $f(x)$.
We use the density estimator in @Pelletier2005-vu. For point $p \in S^1$, the density at $p$ is estimated by
$$
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{r}),
$$
where $d=1$, $r$ is the bandwidth, $\theta_{X_i}(p)$ is the volume density function at sample point $X_i$, and $d_g(p, X_i)$ is the geodesic distance between $p$ and $X_i$ along the manifold $S^1$.

Since the manifold is known ($S^1$), we know the geodesic distance is the arc length between two points on the manifold, which is given by $L(\gamma) = r(\beta - \alpha)$ with $r=1$. Therefore, the geodesic distance is
$$
d_g(p, X_i) = \arccos(p) - \arccos(X_i)
$$
As stated in @Pelletier2005-vu, in terms of *geodesic normal coordinates* at $X_i$, the volume density function $\theta_{X_i}(p)$ equals the determinant of the metric $g$ expressed in these coordinates at the logarithm map $\exp_{X_i}^{-1}(p)$, i.e. $|g(\exp_p^{-1}(x_i))|$.

[When using normal coordinates(orthogonal basis) for tangent vector in $T_{X_i}S^1$, the exponential map has *geodesic normal coordinates*.]

The Riemannian metric $g$ for $S^1$ is given by the inner products on the tangent spaces of $S^1$. For each $X_i \in M$,
$$
g_{X_i}: T_{X_i}M \times T_{X_i}M \rightarrow \pmb{R},
$$
is smooth in a local neighborhood $U$ of $M$.

Then the volume density function $\theta_{X_i}(p)$ is the inner product of the tangent vectors produced by the logarithm map from $X_i$ to $p$, which is given by
$$
\theta_{X_i}(p) = \sqrt{|\cos(d_g(p,q))|} = \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}.
$$
Therefore, the estimator for $p \in S^1$ is given by
$$
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}} K\bigg(\frac{\|\arccos(p) - \arccos(X_i))\|}{r}\bigg),
$$
where the bandwidth $r$ is selected with the plug-in bandwidth selector implemented by the R package *ks*.

### How to derive volume density function $\theta_{X_i}(p)$

#### Example of $S^2$ in @Henry2009-om.

Let $S^2$ be the two-dimensional sphere of radius 1 and $p \in S^2$. Let $v$ and $w$ be vectors such that $\|v\| = \|w\| = 1$ and $\{v, w, p\}$ is the orthonormal basis of $\mathbb{R}^3$. Consider the exponential chart $(U, \Psi)$ induced by the parametrisation $\Psi^{-1}: B_\pi(0) \rightarrow S^2 \backslash \{-p\}$ given by
$$
\Psi^{-1}(s,t) = \cos(r)p + \frac{\sin(r)}{r}(sv+tw),
$$
if $(s,t) \neq (0,0)$ and $\Psi^{-1}(0,0)=p$ where $r=\sqrt{s^2 + t^2}$.
Note that $r=d_g(p,q)$ if $q = \exp_p(sv+tw)$.
Then we take the partial derivatives of $\Psi^{-1}(s,t)$ with respect to the basis of the tangent space, $s$ and $t$, and have that
$$
A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}s}{r}p + \frac{\cos(r)rs^2+t^2\sin(r)}{r^3}v + \frac{(\cos(r)r-\sin(r))st}{r^3}w,
$$
$$
B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}t}{r}p + \frac{(\cos(r)r-\sin(r))st}{r^3}v + \frac{\cos(r)rt^2+\sin(r)s^2}{r^3}w.
$$
The coefficients for $v$ and $w$ gives the local coordinates of two tangent vectors.
Further, since the riemannian metric is defined as the inner product between two tangent vectors, each component of the riemannian metric could be calculated as $g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))$.
Therefore, we obtain that the volume density on the sphere is
$$
\theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2} \\
= \frac{|\sin(d_g(p,q))|}{d_g(p,g)},
$$
for $q \neq p, -p$, and $\theta_p(p) = 1$.

#### $S^1$

Similarly in $S^1$, we could let $\{v, p\}$ be the orthonormal basis of $\mathbb{R}^2$. Then the exponential chart is induced by the parametrization
$$
\Psi^{-1}(s) = \cos(r)p + \frac{\sin(r)}{r}sv,
$$
and $\Psi^{-1}(0) = p$ where $r = s$. Again $r = d_g(p,q)$ if $q = \exp_p(sv)$. Then we have
$$
A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = - \sin(r)p + \cos(r)v.
$$
So the riemannian metric is $g_q = \cos(r)$ and further,
$$
\theta_p(q) = \sqrt{|\cos(r)|} = \sqrt{|\cos(d_g(p,q))|},
$$
where the geodesic distance is $d_g(p,q) = \arccos(p) - \arccos(q)$, and $\theta_p(p) = 1$.
